Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida

### 2022-01-27 revision/review_response.Rmd

P6/L24-39: that is an impressive and extensive list of agencies and partners; any potential drawbacks to so many data streams?

P6/L54-65: is it routine to take water samples directly from the surface? Ultimately, were the authors worried about any potential stratification, particularly of Karenia brevis.

P7/L32-37: why get data from Tampa Airport and the Airfield in St. Petersburg? There is an NWS location not far from Piney Point.

* __Response__: We have updated our weather data to include wind and precipitation from Albert Whitted Field in St. Petersburg and have excluded the precipitation data from Tampa International Airport.  Our initial rationale for including Tampa was a longer period of record.  However, we have revised the historical baseline in the plot to include only years 2006 to 2020, consistent with our comparison of the water quality data with the historical observations.  Also, the NWS location is in Ruskin, Florida and is farther from the bay proper than the location in St. Petersburg.

@@

Page 6, Lines 46-54 - More information is needed on laboratory procedures. In particular, with many different entities running samples and presumably utilizing different analytical approaches for each variable, it is important to know how comparable the data is between programs.

Page 7, Line 22 - Were the breakpoints for K. brevis, or total phytoplankton abundance?

Page 7, Line 37 - What about from ungauged flows?

* __Response__: Our methods for estimating inflows into Tampa Bay are based on methods that are similarly presented in a technical report published by the Tampa Bay Estuary Program in 2012.  This report provides the basis for compliance assessment of our TMDL for reporting to the Florida Department of Environmental Protection and represents our best estimate at hydrologic load inputs in to the bay.  We have provided a citation to this report in the paper.

     Janicki Environmental, Inc., 2012. Development of a screening level tool for estimating annual hydrologic loadings to Tampa Bay (No. 05-12). Tampa Bay Estuary Program, St. Petersburg, Florida.

### 2022-01-27 manu-draft.Rmd
better rain data, link about inflow estimates

Phytoplankton samples included a mix of quantitative (cells/L) and qualitative (presence/absence) samples for major taxa at similar frequency and spatial distribution as the water quality samples. Harmful Algal Bloom (HAB) data for *Karenia brevis* were obtained from event-based monitoring samples from the [FWC-FWRI HAB Monitoring Database](https://myfwc.com/research/redtide/monitoring/database/).  Phytoplankton blooms were described qualitatively as low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L. Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx). Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.  Finally, precipitation data were from Tampa International Airport, inflow estimates to Tampa Bay were basd on summed hydrologic loads of major tributaries from US Geological Survey gaged sites, and wind data were from Albert Whitted Airfield at St. Petersburg, Florida. Additional details of the sampling methods and data sources are provided in supplement.

@@

Bloom concentrations of the red tide species *K. brevis* in 2021 were first observed in Tampa Bay the week of May 23, with concentrations peaking (10$^6$ to 10$^7$ cells/L) by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b). The increase in *K. brevis* from April to July was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10$^5$ cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The two highest concentrations were observed in 1971 (20 million cells/L) and 2021 (17.6 million cells/L), both being over an order of magnitude above the high category.  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (1995 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth in 2021 (Figure S2f, S3f), in addition to the elevated nutrient concentrations from the Piney Point discharge.

>>>>

Phytoplankton samples included a mix of quantitative (cells/L) and qualitative (presence/absence) samples for major taxa at similar frequency and spatial distribution as the water quality samples. Harmful Algal Bloom (HAB) data for *Karenia brevis* were obtained from event-based monitoring samples from the [FWC-FWRI HAB Monitoring Database](https://myfwc.com/research/redtide/monitoring/database/).  Phytoplankton blooms were described qualitatively as low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L. Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx). Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.  Finally, precipitation and wind data were from Albert Whitted Airfield at St. Petersburg, Florida and inflow estimates to Tampa Bay were based on summed hydrologic loads of major tributaries from US Geological Survey gaged sites [similar to @Janicki12]. Additional details of the sampling methods and data sources are provided in supplement.

@@

Bloom concentrations of the red tide species *K. brevis* in 2021 were first observed in Tampa Bay the week of May 23, with concentrations peaking (10$^6$ to 10$^7$ cells/L) by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b). The increase in *K. brevis* from April to July was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10$^5$ cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The two highest concentrations were observed in 1971 (20 million cells/L) and 2021 (17.6 million cells/L), both being over an order of magnitude above the high category.  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (2006 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth in 2021 (Figure S2f, S3f), in addition to the elevated nutrient concentrations from the Piney Point discharge.
